\documentclass[11pt, a4paper, headings=standardclasses]{scrartcl}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=3.5cm]{geometry}
\usepackage[style=authortitle-ticomp, autocite=footnote, backend=biber, dashed=false]{biblatex}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing
\bibliography{Facharbeit}
\begin{document}
\subject{Facharbeit}
\author{Jakob Sinn \\ Submission on February 28 \\ to Mr. Norris \\ \\ \\ Scheeßel}
\title{The trade between Asia and Europe \\ up to the 19th century \\ and its effects on India}
\maketitle
\thispagestyle{empty}
\clearpage
\pagenumbering{Roman}
\begin{singlespace}
\tableofcontents
\clearpage
\end{singlespace}
\section{Introduction}
\pagenumbering{arabic}
\subsection{Summary}
In this research paper, the effects of Asia--Europe trade on India will be examined.

Both the trade routes themselves as well as instances of exchange, cultural or political, are examined in a chronological structure. The spread of Islam and the Mughal empire as well as the first European expeditions around Africa are described. The rise and fall of the East India company and then the Company Raj, and the political and economic changes that accompanied them, are described.

In conclusion, the trade helped introduce Islam to India and influenced its economic structure from the Middle Ages to the 19th century. The colonization of India was motivated by trade, and enabled by economic and political issues that were often exacerbated by European traders.
\subsection{Research question}
My research question: How did the trade between Asia and Europe up to the 19th century affect India politically, economically, and culturally?
\subsection{Methodology}

For research, I primarily relied on \textit{The spice trade of the Roman Empire} by J. Miller and \textit{The rise and fall of the East India Company} by R. Mukherjee, as well as the articles by Paul Freedman. I had access to these books thanks to the Staats- und Universitätsbibliothek Bremen. Several articles from the Encyclopaedia Britannica, available online, provided context, showed the way to other sources, and filled in gaps. Furthermore, resources by the UK Parliament and UNESCO and several other articles were very useful for niche topics.
\section{Trade in Antiquity}
Trade in this time, before the trade movements were consolidated under the peace and security offered by the Islamic conquests of Arabia, moved both on the land route via the middle east and from India over the India ocean to the Red Sea, and from there on to Egypt or the eastern coast of the Mediterranean.\autocite[Chapter 7]{Rome} For both of these routes, the Mediterranean served as the hub to Europe.\footnote{These routes are recorded in the first--century BC anonymous work \emph{Periplus of the Erythraean Sea}.}

\subsection{Traded goods}
While most of the goods traded were spice and other special, often medical, plants, luxuries like silk and gems were also increasingly in demand from a prosperous upper class in the Roman empire.\autocite{RIS}
\paragraph{Spice}
At the time, spices were appreciated not just as enhancers of food, but primarily as medicine\autocite{MST}. The spices came from India, modern-day Indonesia, and China. Cinnamon, for example, is native to Sri Lanka, Myanmar, Bangladesh, and the Malabar Coast of India itself\autocite{Cinnamon}.
\paragraph{Tea}
Recent find of tea in a grave in western China suggest that tea was not only used in China, but exported along the Silk Road before 141 BC.\autocite{npr} Interestingly, the actual word used for tea illustrates the export along the Silk Road: languages nearer the Persian Empire, which acquired tea overland from northern China, where tea is called ch\={a}. Where tea was imported overseas by the Dutch merchants in the 17th and 18th centuries, who loaded the tea in Fujian and Taiwan, tea is called tea, or tee, or th\'{e}, as the people in those ports called tea t\^{e}.
\paragraph{Silk}
Silk had not yet reached the prominence it would in the late Middle Ages, but was still a well-known import via northern China. There are recorded complaints about the desire of Roman wives for silk depleting the silver and gold reserves, demonstrating the trade deficit between west and east.\autocite[Chapter 13]{Rome}
\subsection{Trade routes}
In this age, India was mostly a distribution hub for the spices and other luxury goods from Indonesia and China. Reports of Chinese trading fleets exchanging goods in Indian harbours even made it all the way to ancient Greece and Rome, where they were written down and are still preserved today, for example in the writings of Ptolemy.\autocite[p.~148-150]{Rome}

\paragraph{To India from Indonesia}

Spices were mainly originating from the islands of Indonesia and Malaya, and, over their history, many empires and kingdoms thrived controlling the trade routes between the isles.\autocite{Indonesia}

There was also a direct trade between Indonesia and western Africa. This led to the colonisation of Madagascar by the Austronesians.\autocite{Madagascar}

The stories of their \emph{Raft-men}, who came to India on rudderless rafts to sell the spices and return home with traded goods, travelled along the trading routes to Europe. They are mentioned, for example, by Pliny.\autocite[Chapter 8]{Rome}

\paragraph{From China}

From China, there were three main routes.

The northernmost, known as the \emph{Scythian route}, went from the north of the yellow river, next to the central silk road. In modern-day north-west China, it separated, and made its way north of the Aral and Caspian Sea, passing through Volgograd and ending on the mouth of the river Don on the Black sea, the ancient Greek colony of Tana\"{i}s. This, of course, opened the way to Byzantium via the Black Sea.\autocite[149]{Rome}

Most of the non-spices exported from China, primarily silk, were carried along the \emph{Silk Road}. The route tended to start in northern China, bypassing Tibet, the Himalayas, and thus India, to the north. From there, it led through the Parthian empire\autocite{SilkRome} to Mediterranean ports like Antioch.\autocite[Chapter 7, Maps 2 and 3]{Rome}

Many spices were imported into the southern ports of China like Canton, from where spices and other goods were also exported by sea to modern-day Sri Lanka. Furthermore, goods were also shipped overland from southern China to India.\autocite[Maps 4 and 5]{Rome} The importance of this route must not be underestimated; Chinese records describe Roman trade missions in cities in modern-day northern Vietnam, and Ptolemy tells of a place called \emph{Cattigara}, possibly \emph{Oc Eo}, an archaeological site in Vietnam, where Roman coins have been found.\autocite{OcEo}

\paragraph{India, a centre of trade}
India, while producing many of the goods traded, also served as a trading centre of the Asia-Europe trade. This was due to its geographical advantages with access to Arabia and the Red Sea to the west, as well as the Indian ocean and land connections to China to the east.

In general, trade flowed mostly eastwards, while the recipients in Europe mostly paid in gold coin. Indias position as a trading centre, especially after the emergence of the Roman empire, is demontrated by find of Roman-struck gold coinage from the imperial age.\autocite[100]{curtin_1984}\label{gold}

\paragraph{From India to Europe}

From India, goods were transported by ship via the Red Sea and the Persian Gulf, or by land.\autocite[96]{curtin_1984} Around 100 BC, trade patterns shifted to a direct connection from the Red Sea to India instead of following the coast up to Oman and then entering the open sea.\autocite[98]{curtin_1984} This was due to a better understanding of the wind and weather patterns of the Indian ocean, especially the monsoon.\autocite[97]{curtin_1984}

The Parthian empire, crossed by the overland routes, often acted as an intermediary between the Chinese trade missions and the Roman traders. They often obscured the origins of their goods to prevent being bypassed.\autocite{SilkRome}

\paragraph{Development under the Romans}

During the Roman empire, both diplomacy and war opened up new routes. Ships from Europe even made it to southern India, bringing home a new spice: black pepper. It was even added to meals, remaining the only spice used this way in Antiquity.\autocite{RIS}

This made spice trade even more important in relation to other imports.\autocite{SilkRome, Rome} These efforts also made Alexandria, then part of the Roman empire, one of the foremost trading cities in the world.\autocite{SpiceTrade}
\section{Networks of Trade in the Middle Ages}

Between the beginnings of the Islamic empire in the 7th century and the first Portuguese expedition to India, trading routes only underwent small and gradual changes, but the cultural environments and economic structures they connected and traversed changed dramatically.

\subsection{Changes in Arabia and the middle east}

\paragraph{Beginnings of the Arabic empire}
After the first conquests under Muhammad, the Islamic or Arabian empire was created, stretching from Spain to India\autocite[Section \textit{Achievements}]{Umayyad}. This made the land routes to India and China much more attractive. This era formed the golden age of the silk road, with increasing contact between Europe and China.

\paragraph{The Abbasid caliphate}

After the removal of the previous widely unpopular\autocite{Umayyad} Umayyad caliphate, the Islamic world was ruled by the nominally Shia \emph{Abbasid Caliphate}. The beginning of Abbasid rule in 750 CE was followed by the inauguration of the \emph{House of Wisdom} in Baghdad, marking beginning the Islamic golden age.\autocite{Abbasid}

In this time, the caliphate was shaken by civil wars and separatist dynasties, including remnants of the Umayyads.\autocite[Section \textit{End}]{Umayyad} Nonetheless, this centralised power made travel and trade easier along the overland Silk Road.

\paragraph{Cultural influences on trade}

Muhammad himself was a merchant\autocite{Muhammad}, and Islam was, in the first centuries of its spread, very much a trade-based religion. Even though Arab Muslims conquered vast lands, these campaigns were not at all motivated by religion\autocite[200]{Lapidus}, and active conversion efforts were rare. Instead, many converted to access the Islamic trade networks.\autocite{SilkRoadIslam} In this environment, merchants, especially Arabs, were able to significantly expand their trade networks.

\paragraph{Changing trade routes}

Due to the Islamization of the land routes between the Mediterranean, these land routes became more attractive for Chinese trade missions --- they were now able to sell their goods to the large Islamic trade network. However, Christians were reluctant trade via the Levant, embroiled in conflict by the crusades, to the benefit of Alexandria.\autocite[Paragraph 5]{UN}

\subsection{India}

In the context of this research report, India is defined not strictly along the geographical lines used today, but as a culturally heterogeneous area that includes modern-day India, Pakistan, Bangladesh, and the north of Myanmar.

Spice production remained a major economic factor in India. An example is Cardamom, native to Malabar in south-western India.\autocite[71]{Rome}

\paragraph{Political structure during the Middle Ages}

Between the fall of the Gupta around 450 CE\autocite[Section \textit{Gupta}]{India} and the next significant centralized authority, the Islamic Delhi sultanate, in 1206\autocite{Delhi}, India was ruled mostly by small kingdoms and principalities. These states often exercised informal control over their neighbours, and many smaller kingdoms paid tribute.\autocite{India}

\paragraph{Cultural transfers from the west}

While India was mostly an exporter of ideas during antiquity (Buddhism reached China in this period\autocite{SilkRoadIslam}), the trading began having a noticeable influence on India by introducing new insights in medicine and, most importantly, a new religion.

After the 8th century, muslim traders began to bring Islam with them.\autocite{SilkRoadIslam} Many along the trade routes eagerly converted, and to this day, muslims make up a majority of Pakistanis and a significant minority of Indians.\autocite{India} \label{Islam} Scientists in India were also exposed to new ideas from the centres of research in the Islamic world, especially in medicine.\autocite{medicine}

\subsection{Europe}

Between the fall of the western Roman empire and the first excursions by merchants, the spice trade remained an important part of the economy.\autocite{Yale} This was, especially in the Middle Ages, driven by the high demand for these spices in Europe. 

According to Freedman, ``Spices were ubiquitous in medieval gastronomy''.\autocite[3]{MST} Spice usage expanded significantly in comparison to antiquity, but medicinal use remained important.\autocite{RIS}

\paragraph{Venice}

Up until the exploitation of routes around Africa to access the spices more directly, Venice controlled the trade in the Mediterranean sea. Venetian merchants sold the spice to distributors in Europe, having gained their monopoly in the Chioggia war (1387-81) with Genoa.\autocite{SpiceTrade}

\section{From the discovery of a seaway to India to trading companies}

After 1498, the trade routes discussed earlier were bypassed by rounding Africa. The focus was not political, but mostly commercial; in da Gamas second voyage to India, enough profit was made to finance a third.\autocite[Section \textit{The second voyage}]{Vasco}

Soon, however, colonial powers like the Portuguese and the Dutch established permanent settlements and later colonies, followed by large \emph{companies} and English and French colonizers.

\subsection{Portuguese expeditions and colonies}

In 1498, a Portuguese expedition under the command of \emph{Vasco da Gama} reached India by rounding Africa, completely bypassing the traditional trade routes\autocite{VdG}. This expedition made a lasting and deep impact on the trade between Asia and Europe, ending the dominance of the Islamic world and Venice as way stations and distributors of the spice trade.\autocite{GLO}

\paragraph{The first voyage}

The expeditions were financed by King Manuel I, who had ascended to the throne in 1495. While court politics were certainly a main factor in starting the expedition, the explanation for why da Gama was chosen to lead it, the underlying goals were to break the Muslim and Venetian monopoly on trade with the east.\autocite{Vasco}

The expedition established a \emph{factory}, or trading post, in Calicut (known today as \textit{Kozhikode}). After a conflict with Arab merchants in the city, the factory was destroyed by a mob and about 50 Portuguese are massacred, while anchored Portuguese ships are unable to help.\autocite[Paragraph 6]{kol}

\paragraph{The second voyage}

Da Gamas second voyage was intended as a trading mission and a punitive expedition to target Calicut, in response to the massacre of the factory.\autocite{Vasco}

The fleet massacred hundreds of pilgrims, women and children after capturing a Muslim fleet conducting Hajj. He justified this act of piracy as retaliation for the Calicut massacre,\autocite{bloomberg} but his actions would set the tone for Indo-European relations in the next centuries. This massacre, and many later conflicts, are attributable to religious hatred by the colonizers against Muslims\autocite[382]{FT}

\paragraph{The \textit{Estado da India}}

The first Portuguese fort on Indian soil was constructed in 1503 in the domain of the Raja of Cochin, an ally of the Portuguese.\autocite[383]{FT} The true foundation to the colony, however, was laid in the capture of the island of Goa in 1510. Aimed at fully controlling the spice trade --- end extorting all traders in the India seas --- and aided by the insufficient naval power available to the Indian authorities\autocite[137]{curtin_1984}, this maritime Empire acquired the name \textit{Estado da India}. It captured Malacca in 1511, and Hormuz in the Persian Gulf in 1515, ensuring supremacy on the seas for a century.\autocite[382--383]{FT}

\paragraph{Effects in Europe}

In 1612, English merchants estimated that the imports of pepper, raw silk, cloves, nutmeg, indigo, and mace would have cost \pounds{}1,465,000 imported via the traditional routes. Imported from India, however, the same amount would cost only \pounds{}511,500.\autocite[393]{RF} Obviously, goods brought to Europe via Egypt soon became uncompetitive, to the great disadvantage of Venice.

With this shift in trade routes, and the importation of bullion from the Americas by the Spanish, the western European countries were able to profit off their colonial possessions, shifting the balance of power within Europe significantly.\autocite{EIC}

\subsection{The Mughal empire}

The \emph{Mughal Empire}, occupying northern India from Afghanistan to Assam and Bangladesh, and southern India almost to the coast, was founded in 1526 by Babur. It existed ceremonially up until direct British rule began in 1857.
According to Richards, the new contacts to Europe contributed to the rise of a centralized power in India.\autocite[6]{richards}
\paragraph{Early history}

Babur, a Muslim, came from central Asia and entered India by first establishing himself in Kabul and pushing through the Khyber Pass.
 After a victory in Panipat in 1526, he was able to occupy much of northern India. However, the focus on military campaigns weakened the empire, and his successor Humayun was overthrown and exiled by a rebellion.\autocite[Chapter 1]{richards}

\paragraph{Expansion and begin of decline}

Humayun was able to return in 1555, re-establishing Mughal rule.\autocite[12]{richards} Akbar, who acceded to the throne in 1556 after his fathers death, expanded the empire to encompass the entire Indian subcontinent to the north of the Godavari river. Establishing a new ruling class, a cult around himself, and increasing trade with European companies, Akbar was able to stabilize and expand the Empire.\autocite[16]{richards}

Jahangir, Akbars heir, ruled from 1605 to 1627, and was succeed by Jahan, who erected the Taj Mahal. Under Jahan, however, the financial situation of the empire was under stress due to the splendour of its court.\autocite[Chapter 6]{richards} After he fell ill, his more liberal son Dara Shikoh became regent, but was overthrown by orthodox Muslims and his younger brother Aurangzeb. Even after Jahan recovered, he was declared unfit to rule and imprisoned.\autocite[Chapter 7]{richards}

\paragraph{Aurangzeb}

Aurangzeb ruled from 1658 to 1707\autocite[165]{richards}. His legacy is controversial, with religious orthodoxy and the establishment of sharia on the one hand, but rising political and economic fortunes on the other. He was able to expand the empire to include most of southern Asia.\autocite[Chapter 8]{richards}

\paragraph{Decline}

After Aurangzeb's successor died in 1712, the dynasty was embroiled in infighting and sank in chaos. In 1719 alone, four emperors ruled successively.\autocite[Chapter 12]{richards} How the empire fell so rapidly is debated, with Richards arguing that it was caused by the empires own prosperity allowing the provinces to seek out independence.\autocite{MughalFinances}

Under Muhammad Shah, who was able to rule until his death in 1748, the Mughal empire began to fully collapse. Thereafter, Mughal power was limited at best, with many de facto independent princes still ceremoniously paying homage to the emperor.\autocite{bose}

\subsection{The East India Company}

The East India Company began as a merchant trading house and was only incorporated as a permanent stock company in 1657.\autocite{EIC2} Up until the latter half of the 18th century, the company was not engaged in political processes.

Due to this research papers focus on India, only the East India Company will be discussed in depth.

\paragraph{Founding}

Limited first colonial attempts, mostly focused on spice-producing islands, were soon eclipsed by private enterprises. Groups of merchants divided the significant risks, and the profits, of an expedition to eastern Asia among themselves. Soon, these temporary ventures were officially sanctioned by the government and became stock-issuing companies in the modern sense.

Two of them, the English East India Company and the Dutch United East Indian Company (the VOC, \textit{Vereenigde Oostindische Compagnie}), emerged as the main powers in the spice islands.\autocite{VOC} In the middle of the 17th century, the East India Company changed its focus to India.\autocite[Part I]{EIC}

\paragraph{First posts in India}

Usually, bullion was exported from Britain and exchanged in India for goods.\autocite[Section \textit{The commodity structure of trade}]{FT} After focusing on the spice islands in the first decades of the 17th century, the company established two trading posts, one of whom became Chennai, and another one in Bombay. These posts, or factories, served as bases for trading with Indian merchants and negotiations with the local authorities.\autocite{EIC}

\paragraph{Trade with India}
\label{balance}
According to Chaudhuri, ``It is inconceivable that European trade with India [...] could have been sustained on a large scale without the discovery of American silver-mines.''\autocite[395]{FT} This illustrates the trade balance between India and Europe, a problem resurfacing later with China and the tea trade. In 1621, the East India Company exported \pounds{}200,000 in bullion and coin.\autocite[398]{FT}

In 1613, the East India company first imported \textit{calico}, an Indian cotton cloth, to Europe for sale. While samples had reached Europe earlier, the new bulk transport available with larger ships made the import of these finished goods economically viable.\autocite[400]{FT, Calico}

While mercantilists condemned the use of Indian textiles in Europe, pepper imports peaked in the 1670s\autocite[399]{FT} and textile became the most important export of India. This benefited the East India Company, having already been driven out of the spice trade (less so in pepper) by their Dutch competitors.\autocite[401--407]{FT}

\paragraph{Competitors}

Other European trading companies, namely the Spanish, Dutch, Danish, and french, were able to establish trading posts in India. For example, the Netherlands controlled the coastal areas of Sri Lanka, having won them from the Portuguese at the behest of the king of Kandy.\autocite[Chapter 5]{dutch}

The East India Company was able to expand its advantage and achieve dominance in India with the help of the English, and later British, navy. This naval power was supplanted by a private army consisting of Indian and British troops, and skilled diplomacy.\autocite{ms} The french, the last remaining competitors, were expelled during the seven years war and never regained a stable foothold.\autocite[132-33]{RF} 

The Portuguese were able to retain a small foothold on the western coast of India, centred around Goa. Portuguese India ceased to exist only in 1961, more than three hundred years after its inception.\autocite{PortugueseIndia}

\section{India as a colony}

\subsection{Motives and opportunities for colonization}

As Sen writes, ``the British search for political control was clearly driven by financial interests''\autocite[Section 3]{tnr}. After the Seven Years War, the East India Company was able to rule larger and larger parts of India, directly or via puppet rulers. This system became known as the \textit{Company Raj}.\autocite{parliament}

This development was enabled by the political and cultural disunity in India, and the growing technological advantage of the British over the Indians. Furthermore, many local rulers were willing to accept British influence for economic and political advantages.

\subsection{British military conquests}

\paragraph{Bengal}
The first major military action by the Company was the conquest of Bengal, an area shared today between India and Bangladesh. In 1717, the Company had been granted a \textit{firman} by the Mughal emperor, freeing them from all inland duties in Bengal.\autocite[257]{RF} Siraj-ud-Daula, ruler of Bengal since 1756, wanted to reverse this grant and attempted to force the British out by force. A year later, this led to open war.\autocite[258]{RF}

In the battle of Plassey, the last independent \textit{navab} of Bengal and his french allies were defeated.\footnote{This conflicts was a part of the seven years war.} The British, commanded by colonel Clives, installed a puppet \textit{navab}, Mir Jafar. He acceded to all British demands, and from now of Bengal was de facto a British colony.\autocite[268]{RF}

\paragraph{Other Indian rulers}

After these first colonial conquests, smaller rulers were willing to accept British rule\autocite[Chapter 4, sections 8 and 9]{RF}, and the Marathas remained as the most powerful Indian empire. The British had made arrangements with local rulers to act as a bulwark against Magratha invasion.\autocite[273]{RF}

Soon, internal differences gave the British an opportunity for action against the Magh\-ratas, and in 1803, lord Lake entered Delhi. By the second decade of the nineteenth century, the British had conclusively defeated any opposition in northern India.\autocite[274]{RF}

Expanding British supremacy over central India was not difficult, as local lords were bankrupt and not able to resist British demands. They lost their military power, but remained as nominal rulers receiving pensions from the British.\autocite[276]{RF}

\paragraph{Serious challenges}

Some areas groups proved more resistant to colonization. Only after the first Anglo-Burmese war 1824--1826, the areas of Assam and Manipur became protectorates. Punjab in the north was annexed after the long and bloody Anglo-Sikh wars between 1848 and 1852.\autocite[277]{RF}

\subsection{Increasing government oversight}

As the East India Company was beginning to rule, directly or indirectly, large parts of India, the government in London began to increase regulation of the Company. In return for its continuing monopoly on the Indian trade, the East India Company lent money to the government and promised to represent the interests of the state.\autocite[394]{RF, parliament}

\paragraph{Regulating Act}

But as the Company had to spend more and more money on administering its new possessions, they were unable or unwilling to fulfil many of their obligations.\autocite[395]{RF} By 1773, many in London felt a need to replace the colonies in north America, as their resistance to British rule became apparent. The first step towards establishing the British \textit{Government} as the rulers of India was made when Lord North's Regulating Act passed parliament.\autocite[395]{RF, parliament}

\paragraph{Pitt's India Act 1784}

The private interest of Company officers in India and the shareholders were severely curtailed, and Council and Supreme Court for India were established. After the Fox and North government had broken down over a similar Act, Pitt carried a bill establishing a Board of Control, and almost entirely wrestling power in the Company from the directors. From now on, the East India Company was de facto a subsidiary of the British government.\autocite[396]{RF, parliament}

\subsection{Changes in India's economy and culture}
 According to Mukherjee, ``Not only were the industries of India devitalized in the first phase of the Company's rule and not only was the social life in India led towards retrogression instead of towards progress, but the exploitation by the Company and its servants attacked the very basis of the Indian economy.''\autocite[335]{RF}
 
This process of economic exploitation, forcing many Indians back into agriculture and deindustrializing the country outside of Company \textit{factories}, has been named a contributor to several disastrous famines in India up to the Second World War.\autocite[Section III]{tnr}

Per Rajan, after about 1825, British attitudes towards Indian culture shifted and attempts were made to anglicise. These attempts ended indifference to British rule and gave rise to nationalist movements in India.\autocite[90]{rajan}

\subsection{Introduction of tea}

By the 18th century, tea had become a popular export of China to Europe. However, the Chinese were not interested in any Indian or European product, so they had to be paid in silver. This depleted the silver reserves of Great Britain, and by extension of the whole of Europe.\autocite[214]{Tea} The East India held a monopoly over the tea trade with China up to 1833\autocite{parliament}.

In response, merchants exported opium grown in India, a popular and illegal drug in China.\autocite[215]{Tea} Soon, however, chinese officials cracked down on the smuggling by tipping 21,306 chests of opium into the Pearl River near Canton.\autocite[203]{Tea} Demands for restitution by the drug-smugglers led to the Opium Wars.\autocite{Opium} With the market forced open, poppys were grown at an even larger scale in India, made into Opium, and traded for tea.\autocite[218-219]{Tea}\label{tea}

As an alternative, expeditions were sent to attempt to grow tea in India, grounded in fears of a Chinese trade embargo similar to the one in Japan.\autocite[211]{Tea} Tea seeds were taken from China, and skilled tea growers were enticed, but often prevented from leaving by the Chinese authorities.\autocite[210]{Tea}

\paragraph{Assam Tea}

This matter was of so great importance that a Tea Committee of twelve was formed specifically tasked with cultivating tea in India.\autocite[210]{Tea} In the end, tea was dicovered in Assam in 1823,\autocite[209]{Tea} and the fist shipment of tea from India reached England in 1838. Today, there are 41,000 tea gardens in Assam, covering about 26 million acres.\autocite{roy}\label{assam}

\subsection{Begin of direct rule}

In 1813, the Company's commercial monopoly was abolished, and in 1834 it was prevented from trading at all. Confined to administering India for the government, it served in this role until a revolt broke out in 1857, ending in direct rule by the crown.\autocite{parliament}

\section{Answers to the research question}

\subsection{Political influences}
\paragraph{The Mughal empire}

The Mughal empire, founded by a descendant of Tamerlane\autocite[6]{richards} with an operating base in Kabul, was founded on connections made due to the spice trade. Furthermore, the influx in bullion due to the export of and trade in spices helped keep the empire running. The inland customs duties formed a large share of the revenue generated by the provincial governments.\autocite[257]{RF}

\paragraph{The British}

As examined in the research paper, the trade connections by the East India Company and other European enterprises formed the basis of the colonization of India. The seaway to India was discovered in an expedition motivated by the profits made by others trading with Asia.

During Company rule, the smooth production of goods to be exported to Europe was the main concern of the authorities. Several famines from the late 18th century to WWII have been attributed to British exploitation of India.\autocite{td}

After independence, the Indian political system was in many aspects modelled on that of the British. For example, the Indian parliament possesses both an upper and lower house, modelled after their British counterparts.

\subsection{Cultural influences}
\paragraph{Islam}
As discussed in \ref{Islam}, the trade connections between the Islamic world and India provided a path for the Islamization of the population of many regions of India, most of which are in Pakistan today. The last half-sentence alone betrays the monumental implications this had on India as we know it today.

The Mughals, coming from central Asia, were Muslims and some attempted to shape India according to their beliefs. Often, religious divisions fractured societies in India, culminating in the bloody and haphazard partition of the British Raj into India and Pakistan in 1947. The \textit{Hindustan Times} estimates that about a million were killed and 15 million displaced, and the conflict between the two nuclear powers continues to this day.\autocite[Chapter VI]{guha, partition}

\paragraph{European influences}

As the colonial power for over 150 years, British influences are still very apparent in Indian culture. Cricket, a sport not widely known outside the Anglosphere, is very popular and culturally present in India.\autocite{cricket} The Indian national team, the \textit{Men in Blue}, have won the world championships twice.\autocite{bbc}

English is still widely spoken and often the dominant language in formal contexts. While Hindi is the official language of India, many states are not Hindi-speaking, and English has been retained for many government purposes. The higher courts in India a conducted in English, with Hindi replacing English in some state courts.\autocite{english}

\subsection{Economic Influences}

\paragraph{Profits from trade}

Before the discoveries under da Gama, Indian merchants were able to act as middlemen between the producers from islands to the east and merchants wanting to transport spices to Europe.\autocite{srilanka} The riches this brought to the coasts of southern India is evident from the numerous containers with coins found by archaeologists, as mentioned in section \ref{gold}.

From a European perspective, the trade balance issues described in \ref{balance} were a problem, but for the merchants and rulers of India, they formed an advantage. With this bullion, they were able to buy goods from China and finance lavish courts and keep armies.\autocite[397]{FT} This stabilised many fragile rulers, further weakening Indian resistance to invasion.

\paragraph{Deindustrialisation of India}

In his book, Mukherjee argues at length that the economic policies of the East India Company resulted in a return to agriculture as the main basis of Indian commerce, suppressing the artisans and merchants who had previously developed their trades similarly to their colleagues in Europe.\autocite[Chapter 5, Sections 1 and 2]{RF}

\paragraph{Cash crops: introduction of tea and opium}

With the introduction of opium production in India to trade for tea with the Chinese (Section \ref{tea}), and later tea to be grown for export (Section \ref{assam}), the structure of the rural economies of India were changed. Parallel to criticisms levelled against American policy in Haiti, this shift away from food self-sufficiency and to food bought on the open market probably contributed to the famines under British rule, along with the arable land lost to tea and poppies.

\section{Conclusion and applications}

After searching for the answers to the research question, one fact stood out: how deeply trading relations have influenced India, and how many chapters in Indian history were co-authored by trade.

In a time of increasing global trade, often between nations that differ greatly in politics and culture, this example of how trade is often much more than just exchanging thing can serve as both a sign of caution and an example of the chances offered by international trade.

\clearpage
\begin{singlespace}
\appendix
\pagenumbering{Roman}
\printbibliography
\end{singlespace}

\end{document}